---------------
Author: Brian Rooney
Course: Reproducible Research - Week 2 - Project 1
---------------

# Introduction

This analysis will examine data from a personal activity monitoring device. This device collects the number of steps taken in 5 minute intervals each day, over the months of October and November, 2012 from an anonymous individual.

## Load and Preprocess the Data
```{r dataload,echo=TRUE}
  temp <- tempfile()
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", temp)
  ## Store activity data in "act" dataframe
  act <- read.csv(unz(temp,"activity.csv"))
  unlink(temp)
  
  act$date <- as.Date(as.character(act$date), "%Y-%m-%d")
```

## Total Number of Steps Taken Each Day
```{r totalSteps,echo=TRUE}
## use aggregate function which ignores missing values by default
totalStepsPerDay <- aggregate(steps ~ date,data=act,FUN = sum)
colnames(totalStepsPerDay) <- c("Date","totalSteps")
totalStepsPerDay
```

### Histogram of Total Number of Steps per Day
```{r histSteps,echo=TRUE}
library(ggplot2,warn.conflicts = FALSE,quietly = TRUE)

## Estimate the number of bins and binwidth for the histogram
numObs <- length(totalStepsPerDay$totalSteps)
numBins <- sqrt(numObs)
maxTotalSteps <- max(totalStepsPerDay$totalSteps)
minTotalSteps <- min(totalStepsPerDay$totalSteps)
bw <- round((maxTotalSteps - minTotalSteps)/numBins)

## Plot the Histogram
ggplot(totalStepsPerDay, aes(totalSteps)) +
  geom_histogram(binwidth = bw) +
  xlab("Total Steps per Day") + ylab("Frequency of Days") +
  labs(title = "Histogram of Total Number of Steps per Day")
```

### Summary Statistics
```{r summarystats, echo=TRUE}
## Overall summary statistics
summary(totalStepsPerDay$totalSteps)

## Mean and Median of Total Steps by Day
meanTotalStepsPerDay <- with(act, tapply(steps,as.factor(date),mean))
medianTotalStepsPerDay <- with(act, tapply(steps,as.factor(date),median))

MeanMedian_TotStepsPerDay <- data.frame(Date = names(meanTotalStepsPerDay),
                                        meanTotalSteps = meanTotalStepsPerDay,
                                        medianTotalSteps = medianTotalStepsPerDay,
                                        row.names = NULL)

MeanMedian_TotStepsPerDay
```

## Mean Number of Steps per Interval
```{r avgStepsByInterval,echo=TRUE}
## Calculate mean number of steps for each Interval
actMeanStepsByInterval <- tapply(act$steps,act$interval,mean,na.rm = TRUE)

## Calculate the MAX Number of Steps and the time interval at which this occurs
maxPoint <- max(actMeanStepsByInterval)
x_Coord_maxPoint <- as.integer(names(which(actMeanStepsByInterval == maxPoint)))

## Plot time series of Mean Num Steps by Interval
par(mar = c(5,4,2,1))
plot(names(actMeanStepsByInterval),actMeanStepsByInterval,type = "l" , lwd=2, xlab="5-minute Interval",
     ylab="Mean Number of Steps",main="Mean Number of Steps per 5-minute Interval")
points(x_Coord_maxPoint, maxPoint,pch=8,col="red")
text(1450,206,label = paste("Max Num of Steps = ",round(maxPoint)), col="red", cex=0.75)
text(x=x_Coord_maxPoint-60,y=0,label=x_Coord_maxPoint,col="red",cex=0.75)
abline(v=x_Coord_maxPoint, lty=2,lwd=1,col="red")
```

The maximum mean number steps is `r round(maxPoint)`, occurring at time interval = `r x_Coord_maxPoint`.

## Missing Data
```{r missingData,echo=TRUE}
## Total number of rows with missing data
summary(act)

missingStepsValues <- sum(is.na(act$steps))
```
The total number of missing values is `r missingStepsValues`, which only occurs in the "steps" variable.

### Replace Missing steps values based on mean Steps for the associated 5-min interval
```{r replaceMissing,echo=TRUE}
## Using actMeanStepsByInterval dataframe from above, replace NA values with mean Number of Steps per 5-min Interval.

# Initialize arrays
obsDate <- as.Date(as.character(array()), format = "%Y-%m-%d")
obsInterval <- array()
obsSteps <- array()

## Loop through each observation (obs) of original dataset and if steps = NA for obs, then use obs' interval
## to lookup mean num of steps for interval and replace NA value with mean Steps for the interval.
for(i in 1:nrow(act)) {
  if(is.na(act[i,1])) {
        obsDate[i] <- act[i,2]
        obsInterval[i] <- act[i,3]
        obsSteps[i] <- round(actMeanStepsByInterval[as.integer(names(actMeanStepsByInterval)) == obsInterval[i]][[1]])
  } else {
        obsSteps[i] <- act[i,1]
        obsDate[i] <- act[i,2]
        obsInterval[i] <- act[i,3]
  }
}

## create new dataset with no missing values
cleanAct <- data.frame(obsSteps,obsDate,obsInterval)
```

## Re-calculate the Total Number of Steps Taken Each Day with new cleanAct dataset
```{r newTotalSteps,echo=TRUE}
newTotalStepsPerDay <- aggregate(obsSteps ~ obsDate,data=cleanAct,FUN = sum)
colnames(newTotalStepsPerDay) <- c("obsDate","newTotalSteps")
newTotalStepsPerDay
```

### Compare Histograms of Total Number of Steps per Day
```{r newHistSteps,echo=TRUE}
library(gridExtra,warn.conflicts = FALSE,quietly = TRUE)

## Original Histogram from above - missing data removed
noMissHist <- ggplot(totalStepsPerDay, aes(totalSteps)) +
  geom_histogram(binwidth = bw) +
  ylim(0,30) +
  xlab("Total Steps per Day") + ylab("Frequency of Days") +
  ggtitle("Histogram of Total Number of Steps \nper Day - No Missing Values")

## New Histogram - missing values replaced
replacedMissHist <- ggplot(newTotalStepsPerDay, aes(newTotalSteps)) +
  geom_histogram(binwidth = bw) +
  ylim(0,30) +
  xlab("Total Steps per Day") + ylab("Frequency of Days") +
  ggtitle("Histogram of Total Number of Steps \nper Day - Replaced Missing Values")

grid.arrange(noMissHist,replacedMissHist,ncol = 2)
```


```{r compareStats,echo=TRUE}
## Stats for Activity Data with no missing values
summary(totalStepsPerDay$totalSteps)

## Stats for Activity with missing values replaced
summary(newTotalStepsPerDay$newTotalSteps)
```

The mean of the 2 datasets is exactly the same, while the median and quartiles 1 and 3 differ slightly.
Replacing missing values had no siginficant impact on the data.

### Compare the Total Number of Steps between both datasets
```{r compareTotalSteps,echo=TRUE}
## Show a side-by-side comparison of Total Steps with NAs to Total Steps with NAs imputed
library(dplyr,warn.conflicts = FALSE,quietly = TRUE)
compareTotals <- merge(totalStepsPerDay, newTotalStepsPerDay,by.x = "Date", by.y = "obsDate", all = TRUE)
colnames(compareTotals) <- c("Date","totalStepsWithNAs","totalStepsNAsImputed")
## Grand Total - totalStepsWithNAs
gtWithNAs <- round(sum(totalStepsPerDay$totalSteps,na.rm = TRUE))
## Grand Total - totalStepsNAsImputed
gtReplacedNAs <- round(sum(newTotalStepsPerDay$newTotalSteps))
```

* The Grand Total Number Steps with NAs = `r format(gtWithNAs, scientific=FALSE)`

* The Grand Total Number Steps with NAs replaced = `r format(gtReplacedNAs, scientific=FALSE)`

## Comparison of Mean Total Steps per Interval between Weekdays and Weekends
```{r WeekdaysWeekends,echo=TRUE}
## Prepare data for plot
library(lubridate,warn.conflicts = FALSE,quietly = TRUE)
typeDay <- as.factor(if_else(wday(cleanAct[,2]) %in% (2:6),"weekday","weekend"))
cleanAct <- cbind(cleanAct,typeDay)

weekdayAct <- subset(cleanAct, typeDay == "weekday", select = obsSteps:typeDay)
weekendAct <- subset(cleanAct, typeDay == "weekend", select = obsSteps:typeDay)

weekdayAvgSteps <- with(weekdayAct, tapply(obsSteps, obsInterval, mean))
weekendAvgSteps <- with(weekendAct, tapply(obsSteps, obsInterval, mean))

dfWeekday <- data.frame(interval = as.integer(names(weekdayAvgSteps)),
                        typeDay = "weekday",
                        avgSteps = as.integer(weekdayAvgSteps))

dfWeekend <- data.frame(interval = as.integer(names(weekendAvgSteps)),
                        typeDay = "weekend",
                        avgSteps = as.integer(weekendAvgSteps))

avgNumStepsIntervalDay <- rbind(dfWeekday, dfWeekend)
```

### Plot of Number of Steps per Interval (weekdays vs weekend)
```{r weekDayEndPlot,echo=TRUE}
library(lattice,warn.conflicts = FALSE,quietly = TRUE)
xyplot(avgSteps ~ interval | typeDay, layout= c(1,2),
       data=avgNumStepsIntervalDay, t="l",
       xlab = "interval", ylab = "Number of steps")
```
